% Lorenz curve for defining core habitat classes in suitability models
% P Solymos
% July 2013

Motivation
==========

Species distribution models often model occurrence
as a function of covariates, and predictions
are binarized based on some threshold.
The threshold is usually based on the assessment
of classification accuracy using the observations
and predicted occurrence probabilities (using AUC, ROC, etc.).

With density (habitat suitability) model, the aim is
to quantify ranking of habitats when occupancy is 1, i.e.
habitat suitability given presence. Therefore, a 0/1
classification is not the best approach because the 
crux is with observation >0. Further,
the counts might be from different protocols, when
counts would become non-comparable. So there is a need
for defining a threshold for expected suitability.

Applications include:

* want to map multiple species and shade the maps so that
  those are comparable across species (e.g. peak and 
  valleys has similar meaning)
* want to delineate core habitats (patch) to calculate
  some other metrics (distance to edge, edge contrast,
  patch size, fragmentation, etc).

A common approach is to use a mean/median or quantile based cutoff
or multiple cutoffs for mapping. Applying the same cutoff(s)
would have different meaning for different species. E.g. it would
under predict core habitats for a widely distributed generalist,
and overpredict for a range restricted specialist.

Lorenz curve
============

The Lorenz curve is used to describe income inequality in finance,
and used in diversity calculations (e.g. right tail sum diversity
profiles based on majorization rule), or to describe
size inequality in plant and animal populations. Its use for
density, habitat suitability model seems quite straightforward,
albeit it seems missing from the literature.
The reason for this gap in the literature can be the fact
that fine grained (habitat level) but large scale (province wide)
abundance models are not really widespread.

Here is a simple simulated example with 25 land cover classes
and with a generalist (abundance is not concentrated
to few classes) and specialist species (habitat is concentrated 
in few classes).

The Lorenz curve plots the cumulative distribution of
the metric of interest (here abundance of a species) after sorting
in increasing order, as a function of the cumulative
distribution of the available population (here area of
land cover types).

Inequality
----------

One diagonal connecting (0,0) and (1,1) points is called the
_line of equality_ because the plot follows this line
when abundance in land cover types is proportional to
the availability of those habitats (density is constant).
The lower the line goes from the line of equality,
the more inequality is to be found in the abundance
distribution. The deviation is quantified by the
Gini coefficient, which is the area between the
line of equality and the empirical curve divided by
0.5 (area under the line of equality). A Gini
coefficient of 0 mean perfect equality, while values
close to 1 indicate high inequality.

```{r lorenzcurve}
set.seed(12345)
nc <- 25
av <- rlnorm(nc)
av <- av / sum(av) # availability for land cover classes
x_g <- rgamma(25, 5, 2) # suitability for generalist
x_s <- rgamma(25, 0.5, 2) # suitability for specialist
library(ineq)
lc_g <- Lc(x_g, n=av)
lc_s <- Lc(x_s, n=av)
plot(lc_g, col=4, 
     xlab="cumulative landscape area", 
     ylab="cumulative population size")
lines(lc_s, col=2)
legend("topleft", col=c(4,2), lwd=2, lty=1,
    legend=c("generalist","specialist"), bg="white")
Gini(x_g) # Gini coefficient for generalist
Gini(x_s) # Gini coefficient for specialist
```

Asymmetry
---------

The deviation from the line of equality
and all indices quantifying the inequality (e.g. Gini index)
give information about the overall inequality in the
statistical population.
The other diagonal connecting the points (1,0) and (0,1)
is called the _line of symmetry_. The name refers to
the fact that the Lorenz curve is symmetric around this line
when the point at which the Lorenz curve has a slope equal to 1, sits
on the line of symmetry.
The curve is called asymmetric when when this point is away from 
the line of symmetry. Th Lorenz asymmetry coefficient describes
this and calculated simply as the sum of x and y coordinates
of the point at which slope of the curve is 1 (parallel with
the line of equality). This coordinate can be calculated
from the data or by differentiating a fitted curve to the data points.
The asymmetry coefficient is 1 under symmetry
(point is on the line of symmetry, the sum of the two coordinates equals 1).

```{r asymmetry}
LAC <-
function (x, n = rep(1, length(x))) 
{
    o <- order(x)
    x <- x[o]
    n <- n[o]
    x <- n * x
    p <- cumsum(n)/sum(n)
    L <- cumsum(x)/sum(x)
    p <- c(0, p)
    L <- c(0, L)

    qp <- qnorm(p) 
    qp <- qp[is.finite(qp)]
    qL <- qnorm(L)
    qL <- qL[is.finite(qL)]
    m <- lm(qL ~ qp)
    Lstar <- pnorm(fitted(m))
    dL <- diff(Lstar)/diff(pnorm(qp))
    i <- min(which(dL > 1))
    out <- list(p=p[i], 
         L=L[i], 
         x=x[i],
         n=n[1])
#    dL <- diff(L)/diff(p)
#    m1 <- max(which(dL < 1))
#    m2 <- min(which(dL > 1))
#    out <- list(p=mean(c(p[m1], p[m2])), 
#         L=mean(c(L[m1], L[m2])), 
#         x=mean(c(x[m1-1], x[m2-1])),
#         n=mean(c(n[m1-1], n[m2-1])))
    out$S <- out$L + out$p
    out
}
lac_g <- LAC(x_g, n=av)
lac_s <- LAC(x_s, n=av)
plot(lc_g, col=4, 
     xlab="cumulative landscape area", 
     ylab="cumulative population size")
lines(lc_s, col=2)
abline(1,-1, col=3)
lines(c(0, lac_g$p), c(lac_g$L, lac_g$L), col=4, lty=2)
lines(c(lac_g$p, lac_g$p), c(0, lac_g$L), col=4, lty=2)
lines(c(0, lac_s$p), c(lac_s$L, lac_s$L), col=2, lty=2)
lines(c(lac_s$p, lac_s$p), c(0, lac_s$L), col=2, lty=2)
abline(lac_g$L-lac_g$p, 1, col=4, lty=2)
abline(lac_s$L-lac_s$p, 1, col=2, lty=2)
legend("topleft", col=c(4,2), lwd=2, lty=1,
    legend=c("generalist","specialist"), bg="white")
lac_g$S # asymmetry coefficient for generalist
lac_s$S # asymmetry coefficient for specialist
```

Thresholds
=========

The threshold I propose is defined by this point where
the slope of the Lorenz curve is 1.
This separates low and high suitability areas
in the landscape based on the distribution
of the data, implicitly taking into account
the differences in equality and asymmetry of the data.

The habitat suitability cutoff is the back scaled
y value from the graph. Core habitat is defined
by land cover classes concentrated right of the
x value ($t_1$).

Another threshold could be based on the point
where the line of symmetry intersects with the
Lorenz curve. This point can be found by calculating
the sum of the x and y coordinates and finding the 
point where the sum equals 1 ($t_2$).

These two cutoff values are equal when the curve is symmetric.

```{r barplots}
z_g <- data.frame(x=x_g, av=av)
z_g <- z_g[order(x_g),]
z_g$L <- cumsum(z_g$x) / sum(x_g)
z_g$p <- cumsum(z_g$av)
z_g$Lp <- z_g$L + z_g$p
z_g$t1 <- ifelse(z_g$L > lac_g$L, 1, 0)
z_g$t2 <- 0
z_g$t2[which.min(abs(z_g$Lp-1)):nc] <- 1

z_s <- data.frame(x=x_s, av=av)
z_s <- z_s[order(x_s),]
z_s$L <- cumsum(z_s$x) / sum(x_s)
z_s$p <- cumsum(z_s$av)
z_s$Lp <- z_s$L + z_s$p
z_s$t1 <- ifelse(z_s$L > lac_s$L, 1, 0)
z_s$t2 <- 0
z_s$t2[which.min(abs(z_s$Lp-1)):nc] <- 1

op <- par(mfrow=c(2,2))
barplot(z_g$x, z_g$av, col=c("white", "blue")[z_g$t1+1],
    main="generalist, t1", ylab="suitability", xlab="land cover classes",
    ylim=c(0,max(z_g$x, z_s$x)))
barplot(z_g$x, z_g$av, col=c("white", "blue")[z_g$t2+1],
    main="generalist, t2", ylab="suitability", xlab="land cover classes",
    ylim=c(0,max(z_g$x, z_s$x)))
barplot(z_s$x, z_s$av, col=c("white", "red")[z_s$t1+1],
    main="specialist, t1", ylab="suitability", xlab="land cover classes",
    ylim=c(0,max(z_g$x, z_s$x)))
barplot(z_s$x, z_s$av, col=c("white", "red")[z_s$t2+1],
    main="specialist, t2", ylab="suitability", xlab="land cover classes",
    ylim=c(0,max(z_g$x, z_s$x)))
par(op)
```

Mapping
=======

The Lorenz curve can be based on ordered predicted values as well.
Using one of the characteristic points ($t_1$, $t_2$) one can scale
the suitability values to a common standard.

```{r mapping}
z_g$x1 <- z_g$x / z_g$x[which(z_g$t1==1)[1]]
z_g$x2 <- z_g$x / z_g$x[which(z_g$t2==1)[1]]
z_s$x1 <- z_s$x / z_s$x[which(z_s$t1==1)[1]]
z_s$x2 <- z_s$x / z_s$x[which(z_s$t2==1)[1]]

op <- par(mfrow=c(2,2))
plot(c(0, z_s$p), c(0, z_s$x1), type="l", col=2, lwd=2, main="t1",
     xlab="scaled suitability", ylab="proportion of landscape")
lines(c(0, z_g$p), c(0, z_g$x1), type="l", col=4, lwd=2)
abline(v=z_g$p[which(z_g$t1==1)[1]], lty=2, col=4)
abline(v=z_s$p[which(z_s$t1==1)[1]], lty=2, col=2)
abline(1,0)

plot(c(0, z_s$p), c(0, z_s$x2), type="l", col=2, lwd=2, main="t2",
     xlab="scaled suitability", ylab="proportion of landscape")
lines(c(0, z_g$p), c(0, z_g$x2), type="l", col=4, lwd=2)
abline(v=z_g$p[which(z_g$t2==1)[1]], lty=2, col=4)
abline(v=z_s$p[which(z_s$t2==1)[1]], lty=2, col=2)
abline(1,0)

plot(c(0, z_g$p), c(0, z_g$x), type="l", col=4, lwd=2, main="t1",
     xlab="unscaled suitability", ylab="proportion of landscape")
lines(c(0, z_s$p), c(0, z_s$x), type="l", col=2, lwd=2)
abline(v=z_g$p[which(z_g$t1==1)[1]], lty=2, col=4)
abline(v=z_s$p[which(z_s$t1==1)[1]], lty=2, col=2)

plot(c(0, z_g$p), c(0, z_g$x), type="l", col=4, lwd=2, main="t2",
     xlab="unscaled suitability", ylab="proportion of landscape")
lines(c(0, z_s$p), c(0, z_s$x), type="l", col=2, lwd=2)
abline(v=z_g$p[which(z_g$t2==1)[1]], lty=2, col=4)
abline(v=z_s$p[which(z_s$t2==1)[1]], lty=2, col=2)
par(op)
```

Next steps
==========

1. Apply these ideas on actual species results,
2. determine core habitats and check sensibility
   of results (White-throated Sparrow,
   CAWA, Yellow-rumped Warbler, BTNW, Brown Creeper,
   Lincoln Sparrow),
3. apply threshold in maps.
4. Possibly use rescaled suitability in
   Marxan type prioritization?
