---
title: "Likelihood based optimal partitioning for indicator species analysis"
author: "Peter Solymos, Ermias T. Azeria. et al."
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: no
    number_sections: no
    toc: no
  word_document: default
---

# Motivation

confounding, data types, null distr, all/rank comb

# Workflow

summaries indices plots etc

## Data and model types

list distributions and give examples (table?)

## Special cases

ordered, RS(P)F

## Extensibility

N-mix, mixedm models

## Uncertainty

# HPC

# Data sets

dolina, ARU date/time, ABMI birds forest age

need a simple example to explain the work flow.

need to rewrite the I value part and explain how scaling of
predictors can lead to funny values with identity link, and thus the tanh.

Figure out other metrics: w (model weight), P (reliability): need to figure something out not just for partitions (P), but habitat based frequencies (wplot).

# Introduction

The distributions of organisms on Earth show different levels of aggregation
determined by climate, vegetation,
species interactions, barriers to dispersal, and anthropogenic disturbances
(cit. Humboldt, Whittaker, Rosenzweig).
Species' responses to environmental factors thus allow predictions,
i.e. identify regions or locations where a species can be found
with higher probability, or in higher numbers.
This knowledge of species-environment relationships has been
at the core of human endeavor, e.g. knowing where to find
food, medicinal plants, species used to make tools from or used as
construction material, according to the archaeological (cit) record and
studies in traditional ecological knowledge (cit).
Species with strong environmental associations are often referred to as
differential, character, or indicator species. These species
are used to characterize certain habitats or vegetation types (Botta-Dukat),
indicate naturalness or degradation of ecosystems (McGeoch & Chown),
measure success of habitat restoration,
alerting about critical community thresholds (TITAN),
or indicate the presence of cryptic or rare species (Beals, Indpower, TWINSPAN).

There are three main types of algorithms that are used to quantify the 
degree of association for species. The contingency table based measures (e.g. the
phi-coefficient, or Chi-square metric) that compare
agreement in binary classifications based on some function of
species' abundances and *a priori* classification of the environment (Chytry etc);
Contingency table based methods quantify association as a correlation
measure indicating the magnitude and sign
(-1: maximum avoidance, 0: no association, +1: maximum preference).
The major limitation of these correlation
measures is that the use of binary classification for species observations
is either based on presence/absence (detection/non-detection) data thus
ignoring possibly available abundance information (e.g. >1 counts), or
is based on arbitrary thresholds when binarizing the abundance data
(Tsiripidis, Tichy). Uncertainty in the strength of associations
can be expressed based on
large sample asymptotics or randomization tests, comparing against the
null hypothesis of no association (Agresti, Count data book).

The analysis of variance (ANOVA) based measures compare
between and within group variance (F-ratio) in species abundance given
an *a priori* classification (Jancey, Wildi). The F-ratio
is used to rank species based on the degree of associations, but
it is not explicitly testing the sign of the associations.
Uncertainty in the strength of associations is commonly expressed
using an F-test and corresponding p-value testing the null
hypothesis of equal abundance.
The parametric assumptions of the ANOVA imply normality and
homoskedastic errors, which might not always be satisfied
in most field situations (e.g. using 0/1, biomass, or % cover data).

The third and most widely adopted approach is the IndVal method
that quantifies the concentration of species occurrence and
abundances given an a-priory classification (Dufrene & Legendre).
The IndVal index combines the species' abundance and occurrence information
into a single index which reflects the magnitude of positive
environment-associations (0-1). Uncertainty
in the strength of associations can be quantified based on
bootstrap (DeCaceres), but this approach is not testing against
the null hypothesis of equal expected abundance within the
partitions, because the method is non-parametric.
The p-value for the null hypothesis is based on permutation tests.
Randomization is used to derive the p-value is based on
randomly placing samples or individuals, and this randomization
might not always be meaningful for continuous input data
(e.g. biomass, or % cover).

A common limitation of the available methods is that
assumptions about the distribution (Binomial, Normal)
or type (0/1, counts) of the species data are too restrictive.
As a result, ecologists need to adjust the input data
(binarization, rounding to integers) to meet the needs of the
analysis options. However, non-count or non-normal
data are commonplace, e.g. vegetation studies measuring the
response on ordinal scale, % cover, or biomass.
Besides, none of these approaches are designed to deal with some other
aspects of field data, for example confounding variables,
sample selection bias (presence-only data),
sampling effort differences, or imperfect detection.

In this paper we introduce a general and extensible likelihood-based
framework for indicator species analysis, that we call the opticut approach.
The opticut approach provides a solution to the limitations
of currently available and used options as listed above.
We compare the power of traditional approaches and opticut
to identify an indicator species when there
is true indication in terms of abundance differences among partitions using
simulations. We also show how to assess uncertainty in the strength
of association, and also uncertainty in classification based on
resampling, thus introducing a wider set of tools for statistical inference.
Finally, we illustrate the breadth of situations where the use of opticut
might be advantageous using case studies.
We also provide the opticut R extension package that implements
computationally efficient algorithm for finding indicator species,
and tools for exploring and visualizing the results.


## Methods

### Parametric model

In indicator species analysis we are faced with the
general problem of finding an optimal
partitioning of low vs. high species abundance with respect to
an a-priory stratification.
The observations are usually organized as a community data matrix with
sites or samples as rows ($i = 1, ..., n$) and species as columns.
We describe the theory for a single species only.
The community-wide inference replicates the analyses for each species
independently of one another. $Y_{i}$'s are observations for a single species
from $n$ locations.
$g_{i}$ is a known descriptor of stratification for the location $i$
that can take any of $K$ discrete values ($K > 1$).
Let us denote $z^{(m)}$ as a binary reclassification of $g$ taking values (0, 1).
The superscript $m = 1, ..., M$ indicates a possible combination of binary reclassification. There can also be other site descriptors denoted as $x_{ij}$
taking discrete or continuous values ($j = 1, ..., p$; the number of predictors).

A suitable parametric model describes the
relationship between the observations and the site descriptors, including
the binary partitioning, through the probability density function
$P(Y_{i} = y_{i} \mid z_{i}^{(m)}, x_{ij}, \theta)$
where $\theta$ is the vector of model parameters:
$\theta = (\beta_{0}, \beta_{1}, \alpha_{1}, ..., \alpha_{p})$.
The choice of the parametric model depends on the nature of the
observations. It can be Gaussian, Binomial, Poisson,
ordinal, Beta regression, or zero-inflated models, with a
suitable link function ($f$) for the mean:
$f(\eta_{i}) = \beta_{0}^{(m)} + \beta_{1}^{(m)} z_{i}^{(m)} + \sum_{j=1}^{p} \alpha_{j}^{(m)} x_{ij}$.

The opticut R extension package implements different parametric models.
The default is a Gaussian linear model with an identity link.
Other models include logistic regression (Binomial distribution with
logistic link function), log-linear count models (Poisson, Negative
Binomial, Zero-inflated Poisson and Negative Binomial with logarithmic link).
Ordinal species observations can be analyzed using
the proportional ordered logistic regression (cit MASS). Percent cover data
may be analyzed based on a Beta regression (cit package).
Presence-only data (e.g. museum records, incidental sightings,
most citizen science observations), or animal movement data can be analyzed
based on resource selection (probability) functions (Lele & Keim).
Other custom distributions can also be defined in the R package,
such as linear mixed models, generalized linear mixed models (cit),
occupancy (cit MacKenzie, Lele et al) or N-mixture models (Royle, Solymos).

### Defining the set of partitions

The maximum number of binary partitions based on $K$ levels is
$2^K - 1$ (see DeCaceres), including complements like
(0, 0, 1) and (1, 1, 0) counted as separate
partitions. Such complementary partitions represent identical stratification,
thus should not affect the result of indicator species analysis.
(It does affect the IndVal value for $K>2$ cases.)
When complements are not counted twice, the number of
possible binary partitions of the stratification with $K$ strata (or levels)
is $M = 2^{K-1} - 1$.
Finding all combinations does not require a model or observed responses.
It only takes a classification vector with $K > 1$ partitions.
As a consequence, the partitions can be set up for all species in the
community matrix at once, and used for finding the best partition for each
species.

Blindly fitting a model to all possible partitions is wasteful
use of resources because the number of candidate partitions
grows steeply with increasing $K$. Instead, one can rank the
$K$ partitions based on expected response values
($\mu_{1}, ..., \mu_{k}, ..., \mu_{K}$,
where $\mu_{k}=E[Y_{i} \mid g_{i}=k, x_{ij}=0]$).
This way we have to explore only $M=K-1$ partitions. For $K=3$
we have the following partitions: (1), (1, 2), (1, 2, 3),
where rank 1 is for lowest and 3 is for highest expected abundance.
The ranking varies from species to species, thus
it is not possible to supply the resulting stratification a-priory
for for all species in the community matrix.
There is an overhead of fitting the model to calculate the ranking first.
But computing efficiencies are significant for $K>3$ cases.
For example for $K=10$ the difference is
56-fold (511 all partitions vs. 9 rank based partitions).

$\widehat{\theta^{(m)}}$ is the maximum likelihood estimate (MLE) of the
model parameters given the data and partition $m$,
with corresponding log-likelihood value 
$l(\widehat{\theta^{(m)}}; y) = l^{(m)}$.
Finding MLEs for all $M$ candidate binary partitions
leads to a set of log-likelihood values. One can compare
the log-likelihood values to a null model (no binary partitioning is necessary)
where we fix $\beta_{1} = 0$ leading to the MLE $\widehat{\theta^{(0)}}$
and corresponding log-likelihood value for the null model:
$l(\widehat{\theta^{(0)}}; y) = l^{(0)}$.
The log-likelihood ratio for each candidate partition can be
calculated as $log(LR)^{(m)} = l^{(m)} - l^{(0)}$.
The best supported binary partition is
the model with the highest log-likelihood ratio value (CITATION).

## Indicator value

The indicator value ($I$) for each candidate partition can be calculated
based on expected values using the inverse link function as
$\mu_{0}^{(m)} = f^{-1}(\beta_{0}^{(m)})$ and
$\mu_{1}^{(m)} = f^{-1}(\beta_{0}^{(m)} + \beta_{1}^{(m)})$.
$I^{(m)} = 1 - min(\mu_{0}^{(m)}, \mu_{1}^{(m)}) / max(\mu_{0}^{(m)}, \mu_{1}^{(m)})$.
Where $\mu_{0}^{(m)} = E[Y_{i} \mid z_{i}^{(m)}=0, x_{ij}=0]$ and
$\mu_{1}^{(m)} = E[Y_{i} \mid z_{i}^{(m)}=1, x_{ij}=0]$ are expected values
for the observations given the binary partition $z_{i}^{(m)}$
and at 0 value for all $x_{ij}$.
The strength of association of the species with a given partition is 
determined by $I^{(m)}$.
The value of $I^{(m)}$ is minimal (0) when $\mu_{0}^{(m)} = \mu_{1}^{(m)}$,
whereas $I^{(m)}$ is maximal (1) when one of the expected values
is close to 0 (i.e. the species is absent). This is also referred to as
high fidelity in plant ecology (citation).
The sign of the association equals $sign(\beta_{1}^{(m)})$. 

The ranking of the $M$ candidate models based on 
$I$ and $log(LR)$ is usually not identical
because $I$ only considers the contrast between predicted
abundances in the two strata, whereas $log(LR)$ represents
goodness of fit and is a function of not only the model
parameters but the data as well. $I$ and $log(LR)$ convey
different information that needs to be presented together
for meaningful statistical inference. Ranking of models
should follow $log(LR)$ (best fit) and not $I$ (largest contrast).

!!! TODO: use DeCaceres 1-species example from Oikos paper to
explain the Gaussian opticut approach.

```{r}
## Gaussian example
library(opticut)
(Y <- c(0, 0, 3, 0, 2, 3, 0, 5, 5, 6, 3, 4))
(z <- as.factor(rep(LETTERS[1:3], each=4)))

opticut1(Y, Z=z)
opticut1(ifelse(Y>0,1,0), Z=z, dist="binomial")

print(opticut1(Y, Z=allComb(z)), cut=-Inf)
print(opticut1(ifelse(Y>0,1,0), Z=allComb(z), dist="binomial"), cut=-Inf)

allComb(z)
rankComb(Y, matrix(1, length(Y), 1), z)

set.seed(123)
oc <- opticut(Y ~ 1, strata=z)
uc1 <- uncertainty(oc, 1, type="asymp", B=9999)
#uc2 <- uncertainty(oc, 1, type="boot", B=999)
#uc3 <- uncertainty(oc, 1, type="multi", B=999)

Y <- cbind(Sp1=c(4, 6, 3, 5, 5, 6, 3, 4, 4, 1, 3, 2),
           Sp2=c(0, 0, 0, 0, 1, 0, 0, 1, 5, 2, 3, 4),
           Sp3=c(0, 0, 3, 0, 2, 3, 0, 5, 5, 6, 3, 4))
oc <- opticut(Y ~ 1, strata=z)
plot(oc)
uc1 <- uncertainty(oc, 1, type="asymp", B=9999)
#uc2 <- uncertainty(oc, 1, type="boot", B=999)
#uc3 <- uncertainty(oc, 2, type="multi", B=999)

```

## Quantifying uncertainty

### Model weights

The relative support for the candidate partitions can be measured
by the log-likelihood ratio. A value of >2 means support for
a partition over the null model, a value of >8 indicates strong support
over the null model. Model weights can be calculated as 
$w^{(m)} = \Delta^{(m)} / \sum_{m=1}^{M} \Delta^{(m)}$,
where $\Delta^{(m)} = exp(l^{(m)} - max(l^{(1)}, ..., l^{(M)}))$ (Burnham & Anderson).

### Asymptotic (large sample) inference

Describe how to get asymptotic CIs for mu0, mu1, I (but best partition is fixed). This is of little value, because we rarely test if I for one species is
different from I for another species. Also, significance of differences between
mu0 and mu1 (contrasts) are of little value on their own. But one might want to 
explore the distributions and interpret the results visually.

### Bootstrap based inference

Small sample based inference (CIs can be asymmetric and wider than
asymptotics). Best partition can be fixed (same applies as for asymptotics).
Best partition can be selected for each bootstrap iteration, thus
giving a probabilistic measure of 'mixing'. DeCaceres also mentioned this.

## Simulations

K=2 case: study effect of:

* contrast
* confounding
* noise
* misclassification

K=3 case: recovering mixing proportions using bootstrap

## Case studies

# Results

Confounding effect is fixed by opticut (IX),
others are similar (I0: interpret only opticut, IV: IndVal,
PH: phi coefficient, FR: F-ratio). Red = 1, yellow=0 (proportion passed).

```{r,echo=FALSE}
load("~/Dropbox/collaborations/opticut/opticut-simuls.Rdata")
r6v <- sapply(res6, function(z) rowMeans(z[2,,]>0))
op <- par(mfrow=c(2,3))
rr <- r6v[,vals2$b1==0.9 & vals2$b3==1]
for (i in 1:5)
image(unique(vals$b2), unique(vals$b4), 
    matrix(-rr[i,], length(unique(vals$b2)), length(unique(vals$b4))),
    main=rownames(rr)[i],
    xlab="Confounding", ylab="Misclassification")
par(op)
```

Power (pass/fail) analyzed with Binomial GLM,
see sign & magnitude of effects and deviance table.
b1: contrast, b2: confounding, b3: noise, b4: misclassification
(small value: small effect etc).

```{r,echo=FALSE}
r6c <- sapply(res6, function(z) rowSums(z[2,,]>0))
xx <- data.frame(success=as.numeric(t(r6c)), 
    failure=B-as.numeric(t(r6c)),
    method=rep(rownames(r6c), nrow(vals2)), vals2)
xx$method <- relevel(xx$method, "IX")
mm <- glm(cbind(success, failure) ~ method + (b1 + b2 + b3 + b4), xx, family=binomial)
summary(mm)
av <- anova(mm)
av$Perc <- round(100 * anova(mm)$Deviance / 1041142, 2)
sum(av$Perc, na.rm=TRUE)
av
```

# Discussion

# References

McGeoch MA and Chown SL (1998) Scaling up the value of bioindicators. 
Trends in Ecology and Evolution 13: 46--47.

Podani J and Csanyi B (2010) Detecting indicator species: some extensions of the
INDVAL measure. Ecological Indicators 10: 1119-1124.

Hill MO (1979) TWINSPAN - A FORTRAN Program for Arranging Multivariate Data
in an Ordered Two-way Table by Classification of the Individuals and Attributes.
Ithaca, New York: Section of Ecology and Systematics, CornellUniversity.

Dufrene M and Legendre P (1997) Species assemblages and indicator species: The
need for a flexible asymmetrical approach. Ecological Monographs 67: 345--366.

De Caceres M and Legendre P (2009) Associations between species and groups of
sites: Indices and statistical inference. Ecology 90: 3566--3574.

De Caceres M, Legendre P, and Moretti M (2010) Improving indicator species
analysis by combining groups of sites. Oikos 119: 1674--1684.

SANDER GREENLAND AND HAL MORGENSTERN 1989. 
Ecological Bias, Confounding, and Effect Modification.
International Journal of Epidemiology, 18: 269--274.

Tsiripidis, Ioannis; Bergmeier, Erwin; Fotiadis, Georgios & Dimopoulos, Panayotis
2009. A new algorithm for the determination of differential taxa.
Journal of Vegetation Science 20: 233--240.

Tichy, Lubomir & Chytry, Milan 2006.
Statistical determination of diagnostic species for site groups of unequal size.
Journal of Vegetation Science 17: 809--818.

Z. Botta-Dukat and A. Borhidi 1999.
New objective method for calculating fidelity. Example: the Illyrian beechwoods.
ANNALI DI BOTANICA LVII: 73--90.

Matthew E. Baker and Ryan S. King 2010
A new method for detecting and interpreting biodiversity and ecological community thresholds.
Methods in Ecology and Evolution, 1, 25--37.

O. Wildi and E. Feldmeyer-Christe 2013.
Indicator values (IndVal) mimic ranking by F-ratio in real-world vegetation data.
COMMUNITY ECOLOGY 14(2): 139--143.

E. van der Maarel 1979.
Transformation of Cover-Abundance Values in Phytosociology and Its Effects on Community
Similarity. Vegetatio, Vol. 39: 97--114.

Chytry, Milan; Tichy, Lubomir; Holt, Jason & Botta-Dukat, Zoltan 2002.
Determination of diagnostic species with statistical fidelity measures.
Journal of Vegetation Science 13: 79--90.

KENNETH P. BURNHAM & DAVID R. ANDERSON 2004.
Multimodel Inference - Understanding AIC and BIC in Model Selection.
SOCIOLOGICAL METHODS & RESEARCH, 33: 261--304

Wildi, O. 1989. A new numerical solution to traditional phytosociological
tabular classification. Vegetatio 81: 95-106.

McGeoch, M.A. and Chown, S.L.1998. 
Scaling up the value of bioindicators. 
Trends Ecol. Evol. 13: 46-47.

Jancey, R.C. 1979. Species ordering on a variance criterion. 
Vegetatio 39: 59--63.



